### 2017-10-02 chapters/introduction/physiology.tex
Incorporate Sams suggestions

The flow from one chamber to another is driven
by pressure gradients, that is blood flows from regions with higher
pressure to regions with lower pressure. Consequently, the lowest
pressure is found in the veins entering the heart, and the highest
pressure is found in the left ventricle. Between the different
chambers there are valves that open and close in response to pressure
gradients as illustrated in Figure \ref{fig:heart_anatomy}.

@@

Starting from the outside of the heart we have the \emph{pericardium},
the \emph{epicardium}, the \emph{myocardium} and the \emph{endocardium}. 
The endo- and epicardium are respectively the inner and outer layers of the
myocardium. The myocardium is made up by individual muscle cells, or
myocytes, which branch and join neighboring cells, and form a
complicated fibrous network which is often referred to as a 
\emph{functional syncytium} (meaning that the number of active muscle
fibers cannot be varied from beat to beat). Each myocyte is composed
of bundles of myofibriles, which again contains
\emph{myofilaments}. The myofilaments consist of a repeated pattern
of lines and bands which can be seen as a collection of individual
contractile units called \emph{sarcomeres}. A simple sketch of a
sarcomere is shown in Figure \ref{fig:sarcomere}. 
\begin{figure}[htbp]
  \centering
    \includegraphics[width=1.0\textwidth]{chapters/introduction/figures/Sarcomere.png}
\caption{Illustration of the sarcomere structure.}
\label{fig:sarcomere}
\end{figure}

@@

The heart beats in a cyclic manner, about one beat every second.
The contraction of the heart is triggered by an electrical signal called an action
potential originating from specialized cells in the right atrium. When an action
potential is triggered, calcium is released into the cells. These
calcium ions binds to a complex called troponin C located at the thin
filaments, which then exposes the actin to the myosin head. When
myosin binds to actin we say that a \emph{cross-bridge} is formed
between the thin and thick filaments. A \emph{power stroke} is
triggered after ATP hydrolysis, and the sarcomere shortens. 
The signal propagates through the myocardium and along
specialized electrical highways with high conducting cells.

% One way to describe the cardiac cycle is via the \emph{electrocardiogram}
% (ECG), which is a way of measuring the electrical propagation though
% the heart. Another
One way to describe the cardiac cycle is by means of
the pressure and volume inside the chambers. For example, plotting
the left ventricular (LV) volume on the $x-$axis and the LV pressure on the $y-$axis
provides an intuitive representation of the cardiac cycle known as the
\emph{PV-loop}. In Figure \ref{fig:pv_loop} we show an illustration of
a typical PV-loop. At end-diastole (ED), the mitral valve closes, and
the ventricle contracts against a rising pressure. During this phase,
no blood goes in or out of the ventricle and we call it the
\emph{isovolumic contraction} phase. When the
ventricular pressure exceeds the aortic pressure, the aortic valve
opens and the heart \emph{ejects} blood into the body. When the ventricular
pressure drops below the aortic pressure, the aortic valve closes and
pressure is dropping as the ventricle is relaxing, and we enter the
\emph{isovolumic relaxation} phase. The phase when the ventricle
contracts, from end diastole until the end of ejection is called
\emph{systole}, hence this point in the cycle is also know as
end-systole (ES). Similarly, the phase for which the ventricle
relaxes from the beginning isovolumic relaxation to end diastole, is
called \emph{diastole}. When the pressure drops below the atrial pressure 
the mitral valve opens and blood is sucked into the ventricle. In the
final stage of the filling of the ventricle, the atria contracts and
fills the ventricle before we arrive at end-diastole, and the cycle
repeats. The phase of atrial contraction is also known as atrial
systole.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.85\textwidth]{chapters/introduction/figures/cardiac_cycle.png}
\caption{Showing the relation between pressure and volume in the left
  ventricle during the cardiac cycle. To the left we show the a reduced version of
the classical Wiggers diagram with pressure and volume plotted against
time. To the right we show the pressure volume loop
with volume plotted on the $x-$axis and pressure on the $y-$axis. }
\label{fig:pv_loop}
\end{figure}

@@

Similarly, imagine now that the ventricle is at the end-systolic state,
when the muscle cells are contracting at its most forcefully. Changing
the pressure at this state would also result in a change in volume and
this relationship between pressure and volume  is called the end
systolic pressure volume relationship (\emph{ESPVR}). The two pressure
volume relationships are depicted in Figure \ref{fig:intro_pvr}.

@@

The slope of any pressure volume relationship is called the \emph{elastance},
and a famous model known as the  time varying elastance model, relates
a time varying elastance $E(t)$ to the pressure and volume in the ventricle
\cite{sagawa1977end},  
\begin{align}
  P(t) = E(t)( V(t) - V_0 ).
  \label{eq:time_varying_elastance}
\end{align}
Here $P$ is the endocardial pressure, $V$ the cavity volume, $E$ the
time-varying elastance, and $V_0$ the volume axis intercept.

In fact, the end systolic elastance, $E(t)$ with $t$ being time of end
systole, is considered to be the gold
standard in terms of quantifying myocardial \emph{contractility}.
However, it is rarely used because it is difficult to determine
clinically. What should be noted is that end systolic elastance is
independent of load, and therefore serve as measure of the ability of
the ventricle to do work, i.e myocardial contractility.

@@

One type of heart failure can result from conduction block in the
heart. The action potential travels through these electrical highways
of high conducting cells. One of these highways known as the Bundle of
His splits into two parts, the right and left bundle. The right bundle
activates the right ventricle, while the left bundle activates the
left ventricle. For patients suffering from left bundle branch block,
there is ``road block'' on the highway along the left bundle, meaning
that the left ventricle is activated later than preferable, and the
right ventricle is activated prior to the left
ventricle. This can result in a dyssynchronous contraction, yielding a
reduced pumping effect. A promising treatment for these patients is
called Cardiac Resynchronization Therapy (CRT) \cite{cleland2005effect}.

Imagine a boat with rowers on both sides trying to get the boat moving
forward. If the rowers don’t synchronize their strokes, the boat will
veer from side to side and have a much longer route – if it gets
anywhere at all. To synchronize the rowers, we could add a coxswain
who tells the rowers when to row. We can use the rower analogy to
think of the contractions of the left and right side of the
heart. With the heart, implanting a CRT device is like adding a
coxswain to direct the rowers. The CRT’s goal is to try to
resynchronize the heart so that the left and right ventricles contract
simultaneously. It involves pacing the heart on both the left and
right ventricle so that the heart can contract in a synchronous
way. Therefore CRT are often referred to as biventricular pacing, or
just BiV pacing.

Experiments show that between 30-40\% of the selected patients do
not respond to CRT \cite{daubert20122012}, which is why much research
is put into increasing the responder rate.

>>>>

@@

Starting from the outside of the heart we have the \emph{pericardium},
the \emph{epicardium}, the \emph{myocardium} and the \emph{endocardium}. 
The pericardium is a fibroserous sac that surrounds that heart, and
the endo- and epicardium are respectively the inner and outer layers
of the myocardium. The myocardium is made up by individual muscle cells, or
myocytes, which branch and join neighboring cells, and form a
complicated fibrous network which is often referred to as a 
\emph{functional syncytium} (meaning that the number of active muscle
fibers cannot be varied from beat to beat).

The ventricular muscle fibers orientations varies smoothly through the
wall, and rotates from the endocardium to the epicardium forming a
helical structure \cite{streeter1969fiber}. These muscle fiber are
further organized in laminar sheets, which is separated by gaps called
cleavage planes \cite{legrice1995laminar}.

Each myocyte is composed of bundles of myofibriles, which again contains
\emph{myofilaments}. The myofilaments consist of a repeated pattern
of lines and bands which can be seen as a collection of individual
contractile units called \emph{sarcomeres}. A simple sketch of a
sarcomere is shown in Figure \ref{fig:sarcomere}. 
\begin{figure}[htbp]
  \centering
    \includegraphics[width=1.0\textwidth]{chapters/introduction/figures/Sarcomere.png}
\caption{Illustration of the sarcomere structure.}
\label{fig:sarcomere}
\end{figure}

@@

The heart beats in a cyclic manner, about one beat every second.
The contraction of the heart is triggered by an action
potential originating from specialized cells in the right atrium. When an action
potential is triggered, calcium is released into the cells. These
calcium ions binds to a complex called troponin C located at the thin
filaments, which then exposes the actin to the myosin head. When
myosin binds to actin we say that a \emph{cross-bridge} is formed
between the thin and thick filaments. A \emph{power stroke} is
triggered after ATP hydrolysis, and the sarcomere shortens. 
The signal propagates through the myocardium and along
specialized electrical highways with high conducting cells.

% One way to describe the cardiac cycle is via the \emph{electrocardiogram}
% (ECG), which is a way of measuring the electrical propagation though
% the heart. Another
One way to describe the cardiac cycle is by means of
the pressure and volume inside the individual chambers. For example, plotting
the left ventricular (LV) volume on the $x-$axis and the LV pressure on the $y-$axis
provides an intuitive representation of the cardiac cycle known as the
\emph{PV-loop}. In Figure \ref{fig:pv_loop} we show an illustration of
a typical PV-loop. At end-diastole (ED), the mitral valve closes, and
the ventricle contracts against a rising pressure. During this phase,
no blood goes in or out of the ventricle and we call it the
\emph{isovolumic contraction} phase. When the
ventricular pressure exceeds the aortic pressure, the aortic valve
opens and the heart \emph{ejects} blood into the body. When the ventricular
pressure drops below the aortic pressure, the aortic valve closes and
pressure drops as the ventricle relaxes, and we enter the
\emph{isovolumic relaxation} phase. The phase when the ventricle
contracts, from end diastole until the end of ejection is called
\emph{systole}, hence this point in the cycle is also know as
end-systole (ES). Similarly, the phase for which the ventricle
relaxes from the beginning isovolumic relaxation to end diastole, is
called \emph{diastole}. When the pressure drops below the atrial pressure 
the mitral valve opens and blood is sucked into the ventricle. In the
final stage of the filling of the ventricle, the atria contracts and
fills the ventricle before we arrive at end-diastole, and the cycle
repeats.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.85\textwidth]{chapters/introduction/figures/cardiac_cycle.png}
\caption{Showing the relation between pressure and volume in the left
  ventricle during the cardiac cycle. To the left we show a reduced version of
the classical Wiggers diagram with pressure and volume plotted against
time. To the right we show the pressure volume loop
with volume plotted on the $x-$axis and pressure on the $y-$axis. }
\label{fig:pv_loop}
\end{figure}

@@

Similarly, imagine now that the ventricle is at the end-systolic state,
when the muscle cells are contracting at their most forcefully. Changing
the pressure at this state would also result in a change in volume and
this relationship between pressure and volume  is called the end
systolic pressure volume relationship (\emph{ESPVR}). The two pressure
volume relationships are depicted in Figure \ref{fig:intro_pvr}.

@@

The slope of any pressure volume relationship is called the \emph{elastance},
and a famous model known as the  time varying elastance model, relates
the cavity pressure in the ventricle to the volume and a time varying elastance $E(t)$
\cite{sagawa1977end},  
\begin{align}
  P(t) = E(t)( V(t) - V_0 ).
  \label{eq:time_varying_elastance}
\end{align}
Here $P$ is the endocardial pressure, $V$ the cavity volume, $E$ the
time-varying elastance, and $V_0$ the volume axis intercept.

In fact, the end systolic elastance, $E(t)$ with $t$ being time of end
systole, is considered to be the gold
standard in terms of quantifying myocardial \emph{contractility}.
However, it is rarely used because it is difficult to determine
clinically. What should be noted is that end systolic elastance is
independent of load, and therefore serves as measure of the ability of
the ventricle to do work, i.e myocardial contractility.

@@

One type of heart failure can result from conduction block in the
heart. An electrical wave travels through electrical highways
of high conducting cells. One of these highways known as the Bundle of
His splits into two parts, the right and left bundle. The right bundle
activates the right ventricle, while the left bundle activates the
left ventricle. For patients suffering from left bundle branch block,
there is ``road block'' on the highway along the left bundle, meaning
that the left ventricle is activated later than preferable, and the
right ventricle is activated prior to the left
ventricle. This can result in a dyssynchronous contraction, yielding a
reduced pumping effect. A promising treatment for these patients is
called Cardiac Resynchronization Therapy (CRT) \cite{cleland2005effect}.

Imagine a boat with rowers on both sides trying to get the boat moving
forward. If the rowers don’t synchronize their strokes, the boat will
veer from side to side and have a much longer route – if it gets
anywhere at all. To synchronize the rowers, we could add a coxswain
who tells the rowers when to row. We can use the rower analogy to
think of the contractions of the left and right side of the
heart. With the heart, implanting a CRT device is like adding a
coxswain to direct the rowers. The CRT’s goal is to try to
resynchronize the heart so that the left and right ventricles contract
simultaneously. It involves pacing the heart on both the left and
right ventricle so that the heart can contract in a synchronous
way. Therefore CRT is often referred to as biventricular pacing, or
just BiV pacing. However, experiments show that between 30-40\% of the
selected patients do not respond to CRT \cite{daubert20122012}, which
is why much research is put into increasing the responder rate.
